Data Processing Strategies to Determine Maximum Oxygen Uptake: A Systematic Scoping Review and Experimental Comparison with Guidelines for Reporting

### 2022-11-25 paper.qmd
Adjust to feedback by OJQ

**Results** Most studies use binned time averages to determine V̇O~2max~, with a minority using moving time or moving breath averages. Half of the included studies did not report their data processing strategy and almost all articles failed to provide a rationale for the particular strategy chosen. The processing strategies found in the literature can lead to median differences in V̇O~2max~ of more than 5% (range 0-7%) with considerable variation at the individual level.

@@

Researchers predominantly measure V̇O~2max~ during exercise tests to exhaustion. Among the factors that influence the V̇O~2max~ determination are the exercise test protocol [@buchfuhrer1983 ; @yoon2007; @midgley2008], exercise modality [@myers1991], motivation [@midgley2018], and biological variability [@katch1982]. The measuring method (e.g. mixing chamber, breath-by-breath) and the metabolic cart model used can influence the measured oxygen uptake --- and as such the V̇O~2max~ [@winkert2021; @miles1994]. Due to its inherent variation, the collected raw data data can hardly be interpreted without processing. The same oxygen uptake data generated during an exercise test may result in varying outcomes when processed differently [@astorino2009].

@@

Measured gas exchange data is inherently noisy. Both the biological variability of breathing patterns and the measurement error (as well as irregular breaths such as coughs and swallowing) result in large fluctuation of raw oxygen uptake data on a breath-by-breath basis. To interpret the results the raw data therefore requires some form of processing to obtain a more stable estimate. However, different data processing strategies influence measured parameters of gas exchange [@matthews1987; @johnson1998; @astorino2009; @sell2021]. Midgley et al. @midgley2007 found differences in estimated VO~2max~ values between data processing strategies. As no difference regarding the reliability of the different approaches was found, no optimal strategy exists per se. Accordingly, when comparing outcomes within and across studies obviously a consistent processing strategy should be used. Based on data in sedentary and moderately trained individuals, Martin-Rincon et al. @martin-rincon2019 provided linear-log equations to compare mean V̇O~2max~ values derived from processing strategies with differing averaging interval lengths. The results showed that reliability of V̇O~2max~ was higher in trained individuals, but not influenced by the data processing. Furthermore, there is inconclusive evidence on whether the influence of different processing strategies interacts with different exercise protocols [@hill2003; @scheadler2017].

Researchers have proposed a variety of calculation intervals and computational methods to process oxygen uptake data [@howley1995; @robergs2003; @ats/accp2003; @weir2004; @robergs2010] and calls to standardize processing strategies are frequent [@scheadler2017; @myers1990; @hill2003; @johnson1998]. In light of the influence on outcome variables, many articles highlighted the need to report processing strategies in research [@myers1990; @astorino2009; @scheadler2017; @johnson1998]. Midgley et al. @midgley2007 were the first to evaluate reported data processing strategies for breath-by-breath analyses in selected journals. They found that almost all studies reporting the use of binned time averaging, with only 1 in 117 using a moving time and a moving breath average, respectively. One third of the studies did not describe their processing method at all. In our article we will provide a systematic analysis of the reporting and usage of data processing strategies in recent literature.

Robergs et al. @robergs2010 argued that to investigate the current state of data processing strategies, two possible approaches are *"(i) a summary of published research, and (ii) a survey circulated via the Internet to as many exercise physiologists as possible"*. They chose the latter approach with a total of 75 respondents, who reported a great variety in data processing strategies. Most researchers reported the use of binned time averages over 30 or 60 seconds. Surprisingly, about half of the respondents admitted that their data processing strategy was rather chosen due to subjective factors as opposed to objective criteria [@robergs2010]. The present work therefore aims to investigate to what extent reporting and processing practices have followed the recommendations put forth by @midgley2007 and @robergs2003.

Selected data processing strategies have been extensively compared in the literature. Due to the absence of a systematic mapping of current practices, these studies however lacked the reasoning of which strategies to compare. Many studies compared different averaging intervals, but not averaging types (e.g. moving breath vs. binned time) [@astorino2009; @midgley2007]. Martin-Rincon et al. @martin-rincon2019 provided formulas for comparing data processing strategies by investigating a data set of sedentary individuals and recreational athletes, using two different metabolic carts. Therefore in this work motivation and measurement devices may have interacted with the influence of processing strategies. Most comparisons only report mean differences between strategies [@astorino2009; @midgley2007]. No research has yet compared a variety of systematically derived strategies among a group of trained individuals using a standardized measurement set-up.

Differences in estimated V̇O~2max~ values due to variation in data processing can have serious implications in practice. For example, the assessment of longitudinal data from athletes which participated in diagnostics using differing data processing approaches becomes problematic or the pooling of results across studies in meta-analyses [@martin-rincon2019]. Data processing strategies directly affect the occurrence of a plateau in oxygen uptake, the primary criterion for V̇O~2max~ [@astorino2009]. Crucially, in situations where individuals are classified by their V̇O~2max~ --- for example when describing the training status of a study population [@pauw2013; @decroix2016] or evaluating patients for a heart transplantation [@mancini1991] --- differing processing strategies can lead to misclassifications [@johnson1998].

@@

```{r}
#| echo: false
#| fig-cap: "Total durations of the calculation interval of V̇O~2max~ in the reviewed studies. ROBERT SCHLÄGT VOR HIER LIEBER EINE TABELLE ZU NEHMEN. ICH BIN FÜR DIE GRAFIK. WAS MEINST DU?."
#| label: fig-duration
#| out-width: '90%'

@@

```{r}
#| echo: false
#| fig-cap: "V̇O~2max~ varies by data processing strategy. Values are expressed relative to the V̇O~2max~ from a 30-second binned average — the most common strategy as determined by the review. Solid lines display the median, the shaded areas mark the interval between 10th and 90th percentile. Using moving average leads to systematically higher V̇O~2max~ values compared to binned time averages. Changing the averaging interval (in seconds or breaths) can lead to median changes in V̇O~2max~ as large as 5%. XLABEL GEÄNDERT IN 'DURATION' STATT 'PARAMETER'. OKAY?"
#| fig-scap: "Comparison of different data processing strategies."
#| label: fig-comparison
#| out-width: '90%'

knitr::include_graphics("./plots/comparison_xlabel.png")
```

@@

One possible reason for the prevalence of binned averages data processing approaches may be due to limitations in analysis software used. The results show that most researchers use the vendor software of the metabolic cart's manufacturer. These software may by default output binned time averages instead of raw breath-by-breath data. Moreover, further processing (e.g., interpolation, moving averages) may require the use of additional software. This may also explain why digital filtering has --- despite recommended by Robergs et al. @robergs2010 --- not been used in a single study reviewed here: standard distributions of common data analysis software (e.g., Microsoft Excel) lack the capability to perform such operations. The common practices of data processing are likely to be both attributed to a lack of awareness and a lack of adequate software solutions.

@@

The different data processing strategies found in the literature directly bias the V̇O~2max~ determination (see @fig-comparison), and as such influence the classification of individuals, the evaluation of training success, and the assessment of V̇O~2max~ attainment. In accordance with previous findings [@johnson1998; @astorino2009; @sell2021] longer calculation intervals lead to lower V̇O~2max~ values (see @fig-comparison). The analyzed data shows median differences up to 5-7% between processing strategies, which is in accordance with previous research [@martin-rincon2019]. Some studies reported even higher mean differences of up to 20% [@myers1990], but only when including raw breath-by-breath data in the comparison. The evaluation of unprocessed raw data for its maximum is highly erroneous and as such is not performed in research (see @fig-duration); so there is no reason to compare it to other strategies. While previous research was often conducted in sedentary or recreationally trained individuals, we now present evidence, that a similar effect of data processing strategies on V̇O~2max~ exists in trained athletes. Since training interventions in trained athletes typically show improvements in V̇O~2max~ in the range of 0-6% [@parmar2021; @midgley2007a], variation caused by differing data processing strategies can bias the evaluation of their success.

@@

The exact impact of data processing strategies on the V̇O~2max~ is highly individual. Most research did present only comparisons of mean values, with results in accordance with those found here [@martin-rincon2019]. At the individual level, data processing strategies may impact V̇O~2max~ values with varying magnitudes. For example, for 10% of the investigated athletes a binned time average of 5 seconds leads to a V̇O~2max~ \<3% higher than by a 30-second average, while for another 10% the V̇O~2max~ was \>6% higher (see @fig-comparison). Current values reported and equation derived compare strategies on a group level [@martin-rincon2019], which improves comparison of group results for meta-analyses or group classifications. However, at the individual level these equations can only be applied with a large margin of error. Changes in the impact of different data processing strategies on V̇O~2max~ range from 1-2% in some individuals to more than 10% in others. Hence when evaluating V̇O~2max~ data from different tests in a single individual obtained by using different processing methods, there is no way to accurately compare these values even when the processing strategies are reported. While the comparison of V̇O~2max~ from different processing strategies require their reporting for an sufficient analysis on a group level, on the individual level the raw data from each test is required.

@@

To compare and evaluate V̇O~2max~ values from different studies, knowledge of the underlying data processing strategies is crucial. Our review demonstrates that almost half of the studies measuring V̇O~2max~ did not describe their processing strategy. Other aspects of the data processing, such as outlier filtering or the rationale for the chosen procedure were only in rare instances reported (see @tbl-reporting). @tbl-recommendation lists seven items that should be reported to provide sufficient information on the data processing strategy used to determine V̇O~2max~. These items may be reported in form of a checklist, as an in-text enumeration or in sentence form. An example paragraph containing all the relevant information for the original data presented in this paper [@quittmann_unpubl_a; @quittmann_unpubl_b] would be:

@@

Note that the correct reporting of an exercise test to determine V̇O~2max~ requires more information than those on data processing. Further aspects to be reported include, but are not limited to, the study population, the exercise protocol, the device calibration, and criteria to stop the test. In cases where journals endorse word limits on articles, this reporting --- including the reporting on data processing strategies --- may be included in supplementary files. The correct and detailed reporting of data processing strategies, as well as other test characteristics, is crucial for interpreting presented V̇O~2max~ values.

The results of the present work suggest that comprehensive reporting facilitates approximate comparisons of V̇O~2max~ data on a group level derived using different data processing strategies. But on an individual level and for a precise comparison, reporting may not be sufficient, as differences between data processing strategies vary between individuals and are potentially influenced by training status. Sharing of the raw gas exchange data can solve this challenge, as it allows any researcher to recalculate the V̇O~2max~ using their preferred data processing strategy. Most raw gas exchange data files are structured in a simple way, which allows to easily remove any personal information (if this had not been done in the metabolic cart's system before). In terms of reproducibility of V̇O~2max~ determination, a way even superior to correct reporting and raw data sharing is to additionally share data analysis code. This allows any researcher to independently reproduce the V̇O~2max~ processing conducted within a study, but requires the data analysis to take place in a programming (or at least code generating) environment. Such programs for the purpose of analyzing gas exchange data exist as free open-source software [@maturana2022; @nolte2022].

@@

Due to the sheer quantity of the publications investigating V̇O~2max~, it was not possible to perform an exhaustive review of the literature. The scoping review therefore relies on a random sample which not necessarily captures exact trends of the literature. However, efforts were made --- such as random sampling and systematic article exclusions --- to ensure the sample to be representative. Notable, almost half of the studies did not report their data processing strategy at all. The data processing strategies used in the literature could only be investigated when studies reported them.

Ambiguities in the reporting of the investigated studies may impact analysis results. For example some studies using long binned averages (e.g., 60 seconds) may have in fact been using multiple binned averages of shorter duration (e.g., 4x15 second), without describing this correctly. Moreover the exact definitions for building binned averages vary within the literature: While most studies define the binning periods from the beginning of the exercise, some may define them from the endpoint. Additionally some studies reviewed did not define the maximum bin, but a pre-set binned average period as their V̇O~2max~ (for example the last bin, regardless of its value). In situations where the maximum in oxygen uptake is reached considerably before exhaustion (i.e., a long plateau in oxygen uptake exists) this may lead to different results than a traditional binned average processing. We did not separately consider such sub-categories of data processing strategies, as they may not be very common and are often hard to precisely investigate due to ambiguities in their reporting.

@@

>>>>

**Results** Half of the included studies did not report their data processing strategy and almost all articles failed to provide a rationale for the particular strategy chosen. Most studies use binned time averages to determine V̇O~2max~, with a minority using moving time or moving breath averages. The processing strategies found in the literature can lead to median differences in V̇O~2max~ of more than 5% (range 0-7%) with considerable variation at the individual level.

@@

Researchers predominantly measure V̇O~2max~ during exercise tests to exhaustion. Among the factors that influence the V̇O~2max~ determination are the exercise test protocol [@buchfuhrer1983 ; @yoon2007; @midgley2008], exercise modality [@myers1991], motivation [@midgley2018], and biological variability [@katch1982]. Even the measuring method (e.g. mixing chamber, breath-by-breath) and the metabolic cart model used can influence the measured oxygen uptake --- and as such the V̇O~2max~ [@winkert2021; @miles1994]. Due to its inherent variation, the collected raw data can hardly be interpreted without processing. The same oxygen uptake data generated during an exercise test may result in varying outcomes when processed differently [@astorino2009], which can have serious implications in practice [@johnson1998].

@@

Measured gas exchange data is inherently noisy. Both the biological variability of breathing patterns and the measurement error (as well as irregular breaths such as coughs and swallowing) result in large fluctuation of raw oxygen uptake data on a breath-by-breath basis. Therefore, the raw data require some form of processing to obtain a more stable estimate. However, different data processing strategies influence measured parameters of gas exchange [@matthews1987; @johnson1998; @astorino2009; @sell2021]. Midgley et al. @midgley2007 found differences in estimated V̇O~2max~ values between data processing strategies. As no difference regarding the reliability of the different approaches was found, no optimal strategy exists per se. Accordingly, when comparing outcomes within and across studies obviously a consistent processing strategy should be used. Based on data in sedentary and moderately trained individuals, Martin-Rincon et al. @martin-rincon2019 provided linear-log equations to compare mean V̇O~2max~ values derived from processing strategies with differing averaging interval lengths. The results showed that the reliability of V̇O~2max~ was higher in trained individuals, but not influenced by data processing. Furthermore, there is inconclusive evidence on whether the influence of different processing strategies interacts with different exercise protocols [@hill2003; @scheadler2017].

Researchers have proposed a variety of calculation intervals and computational methods to process oxygen uptake data [@howley1995; @robergs2003; @ats/accp2003; @weir2004; @robergs2010] and calls to standardize processing strategies are frequent [@scheadler2017; @myers1990; @hill2003; @johnson1998]. In light of the influence on outcome variables, many articles highlighted the need to report processing strategies in research [@myers1990; @astorino2009; @scheadler2017; @johnson1998]. Midgley et al. @midgley2007 were the first to evaluate reported data processing strategies for breath-by-breath analyses in selected journals. They found that almost all studies reporting the use of binned time averaging, with only 1 in 117 using a moving time and a moving breath average, respectively. One third of the studies did not describe their processing method at all.

Robergs et al. @robergs2010 argued that to investigate the current state of data processing strategies, two possible approaches are *"(i) a summary of published research, and (ii) a survey circulated via the Internet to as many exercise physiologists as possible"*. They chose the latter approach with a total of 75 respondents, who reported a great variety in data processing strategies. Most researchers reported the use of binned time averages over 30 or 60 seconds. Surprisingly, about half of the respondents admitted that their data processing strategy was rather chosen due to subjective factors as opposed to objective criteria [@robergs2010]. The present work therefore aims to investigate to what extent reporting and processing practices have followed the recommendations put forth by Midgley et al. [@midgley2007] and Robergs et al. [@robergs2003].

Selected data processing strategies have been extensively compared in the literature. Due to the absence of a systematic mapping of current practices, these studies however lacked the reasoning of which strategies to compare. Many studies compared different averaging intervals, but not averaging types (e.g. moving breath vs. binned time) [@astorino2009; @midgley2007]. Martin-Rincon et al. @martin-rincon2019 provided formulas for comparing data processing strategies by investigating a data set of sedentary individuals and recreational athletes, using two different metabolic carts. Therefore in their work motivation and measurement devices may have interacted with the influence of processing strategies. Most comparisons only report mean differences between strategies [@astorino2009; @midgley2007]. No research has yet compared a variety of systematically derived strategies among a group of trained individuals using a standardized measurement set-up.

Differences in estimated V̇O~2max~ values due to variation in data processing can have serious implications in practice. For example, the assessment of longitudinal data from athletes which participated in diagnostics using differing data processing approaches becomes problematic. The same applies to the pooling of results across studies in meta-analyses [@martin-rincon2019]. Data processing strategies directly affect the occurrence of a plateau in oxygen uptake, the primary criterion for V̇O~2max~ [@astorino2009]. Crucially, in situations where individuals are classified by their V̇O~2max~ --- for example when describing the training status of a study population [@pauw2013; @decroix2016] or evaluating patients for a heart transplantation [@mancini1991] --- differing processing strategies can lead to misclassifications [@johnson1998].

@@

```{r}
#| echo: false
#| fig-cap: "Total durations of the calculation interval of V̇O~2max~ in the reviewed studies. STILL UP TO DEBATE WHETHER RATHER PLOT (VERTICAL OR HORIZONTAL BARS?) OR TABLE."
#| label: fig-duration
#| out-width: '90%'

@@

```{r}
#| echo: false
#| fig-cap: "V̇O~2max~ varies by data processing strategy. Values are expressed relative to the V̇O~2max~ from a 30-second binned average — the most common strategy as determined by the review. Solid lines display the median, the shaded areas mark the interval between 10th and 90th percentile. Using moving average leads to systematically higher V̇O~2max~ values compared to binned time averages. Changing the averaging interval (in seconds or breaths) can lead to median changes in V̇O~2max~ as large as 5%."
#| fig-scap: "Comparison of different data processing strategies."
#| label: fig-comparison
#| out-width: '90%'

# keeping 'Parameter' as x label, as 'Duration' may be improper
# when describing breath-based averages.
knitr::include_graphics("./plots/comparison.png")
```

@@

One possible reason for the prevalence of binned averages data processing approaches may be due to limitations in analysis software used. The results show that most researchers use the vendor software of the metabolic cart's manufacturer. These software products may by default output binned time averages instead of raw breath-by-breath data. Moreover, further processing (e.g., interpolation, moving averages) may require the use of additional software. This may also explain why digital filtering has --- despite recommended by Robergs et al. @robergs2010 --- not been used in a single study reviewed here: standard distributions of common data analysis software (e.g., Microsoft Excel) lack the capability to perform such operations. The common practices of data processing are likely to be attributed to both: a lack of awareness and adequate software solutions.

@@

The different data processing strategies found in the literature directly bias the V̇O~2max~ determination (see @fig-comparison), and as such influence the classification of individuals, the evaluation of training success, and the assessment of V̇O~2max~ attainment. In accordance with previous findings [@johnson1998; @astorino2009; @sell2021] longer calculation intervals lead to lower V̇O~2max~ values (see @fig-comparison). The analyzed data shows median differences up to 5-7% between processing strategies, which is in accordance with previous research [@martin-rincon2019]. Some studies reported even higher mean differences of up to 20% [@myers1990], but only when including raw breath-by-breath data in the comparison. The evaluation of unprocessed raw data for its maximum is highly erroneous and as such is not performed in research (see @fig-duration); so there is no reason to compare it to other strategies. While previous research was often conducted in sedentary or recreationally trained individuals, we now present evidence, that a similar effect of data processing strategies on V̇O~2max~ exists in trained athletes. Since training interventions in trained athletes typically show improvements in V̇O~2max~ in the range of 0-6% [@parmar2021; @midgley2007a], variation caused by differing data processing strategies is in the same range and can bias the evaluation of their success.

@@

The exact impact of data processing strategies on the V̇O~2max~ is highly individual. Most research did present only comparisons of mean values, with results in accordance with those found here [@martin-rincon2019]. Data processing strategies may impact V̇O~2max~ values with varying magnitudes at the individual level. For example, for 10% of the investigated athletes a binned time average of 5 seconds leads to a V̇O~2max~ \<3% higher than by a 30-second average, while for another 10% of the investigated athletes the V̇O~2max~ was \>6% higher (see @fig-comparison). Current values reported and equation derived compare strategies on a group level [@martin-rincon2019], which improves comparison of group results for meta-analyses or group classifications. However, at the individual level these equations can only be applied with a large margin of error. Changes in the impact of different data processing strategies on V̇O~2max~ range from 1-2% in some individuals to more than 10% in others. Hence when evaluating V̇O~2max~ data from different tests in a single individual obtained by using different processing methods, there is no way to accurately compare these values even when the processing strategies are reported. While the comparison of V̇O~2max~ from different processing strategies require their reporting for an sufficient analysis on a group level, the raw data from each test is required on the individual level.

@@

To compare and evaluate V̇O~2max~ values from different studies, knowledge of the underlying data processing strategies is crucial. Our review demonstrates that almost half of the studies measuring V̇O~2max~ did not describe their processing strategy. Other aspects of the data processing, such as outlier filtering or the rationale for the chosen procedure were only in rare instances reported (see @tbl-reporting). @tbl-recommendation lists seven items that should be reported to provide sufficient information on the data processing strategy used to determine V̇O~2max~. These items may be reported in form of a checklist, as an in-text enumeration or in sentence form. An example paragraph containing all the relevant information for the original data presented in this paper [@quittmann_accepted; @schwarz2022] would be:

@@

Note that the correct reporting of an exercise test to determine V̇O~2max~ requires more information than those on data processing. Further aspects to be reported include, but are not limited to: the study population, exercise protocol, device calibration, and criteria to terminate the test. In cases where journals endorse word limits on articles, this reporting --- including the reporting on data processing strategies --- may be included in supplementary files. The correct and detailed reporting of data processing strategies, as well as other test characteristics, is crucial for interpreting presented V̇O~2max~ values.

The results of the present work suggest that comprehensive reporting facilitates approximate comparisons of V̇O~2max~ data on a group level derived using different data processing strategies. However, on an individual level and for a precise comparison, reporting may not be sufficient, as differences between data processing strategies vary between individuals and are potentially influenced by training status. Sharing of the raw gas exchange data can solve this challenge, as it allows any researcher to recalculate the V̇O~2max~ using their preferred data processing strategy. Most raw gas exchange data files are structured in a simple way, which allows to easily remove any personal information (if this had not been done in the metabolic cart's system before). In terms of reproducibility of V̇O~2max~ determination, sharing the raw data as well as the data analysis code seems to be an even superior way. This requires the data analysis to take place in a programming (or at least code generating) environment. Such programs for the purpose of analyzing gas exchange data exist as free open-source software [@maturana2022; @nolte2022].

@@

Due to the sheer quantity of the publications investigating V̇O~2max~, it was not possible to perform an exhaustive review of all articles. The scoping review therefore relies on a random sample which not necessarily captures exact trends of the literature. However, efforts were made --- such as random sampling and systematic article exclusions --- to ensure the sample to be representative. Notable, almost half of the studies did not report their data processing strategy at all. The data processing strategies used in the literature could only be investigated when studies reported them.

Ambiguities in the reporting of the investigated studies may impact analysis results. For example some studies using long binned averages (e.g. 60 seconds) may have in fact been using multiple binned averages of shorter duration (e.g. 4×15 second), without describing this correctly. Moreover, the exact definitions for building binned averages vary within the literature. While most studies define the binning periods from the beginning of the exercise, some may define them from the endpoint. Additionally some of the included studies did not define the maximum bin, but a pre-set binned average period as their V̇O~2max~ (for example the last bin, regardless of its value). In situations where the maximum in oxygen uptake is reached considerably before exhaustion (i.e., a long plateau in oxygen uptake exists) this may lead to different results than a traditional binned average processing. We did not separately consider such sub-categories of data processing strategies, as they may not be very common and are often hard to investigate precisely due to ambiguities in their reporting.

@@

DISCUSS IF WE SHOULD CHANGE THE ORDER OF RECOMMENDATIONS/GUIDELINES AND LIMITATIONS.
